%!TEX root = ../template.tex
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% acknowledgements.tex
%% NOVA thesis document file
%%
%% Text with acknowledgements
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\typeout{NT FILE acknowledgements.tex}%

\begin{ntacknowledgements}

This thesis represents much more than a year dedicated to this project and its writing. It marks an important period of growth. A period in which I developed multidisciplinary skills by immersion in a fast-paced consulting environment of a multinational company. I would therefore like to express my sincere gratitude to all those who have supported and guided me throughout my journey. 

First, I am deeply grateful to Cecilia, Filippo, and Sarah for believing in me and for offering me the unique opportunity to join such a talented team. A special thank you goes to Filippo for sharing his knowledge and for making the entire journey enjoyable and rewarding, even though the most challenging moments. His guidance, encouragement, and constant support were essential. My sincere thanks also go to all the members of the Discovery and Translational Science Team. Each one, without exception, contributed significantly to my learning process, enriching my experience.

Finally, I am infinitely grateful to my family and friends for their encouragement and unconditional support throughout all the challenges, as well as for the joy shared in all my achievements and successes.


\end{ntacknowledgements}